---
title: "metaGLMM-examples"
author: "Keisuke Hanada"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{metaGLMM-examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## Introduction
The metaGLMM package extends traditional random-effects meta-analysis by embedding it within a generalized linear mixed-effects model framework, thereby enabling covariate adjustment and support for non-normal responses. This methodology was introduced in Hanada (2025); please be sure to cite that paper when using this package.

- Hanada, K. (2025). Working paper. *arXiv*, xxxx.


In this vignette, we demonstrate how to implement the proposed method using the data reanalysis examples presented in Hanada (2025).




## Loading libraries

Set up of Our package and seed value.
```{r setup, message=FALSE}
set.seed(1234)

library(dplyr)
library(ggplot2)

library(metaGLMM) # our package
```


## Example 1: Binary distributed outcome

### Data Preparation

```{r example1, eval=FALSE}
data(chu2020, package="metaGLMM")
dat <- chu2020
dat
#>     Study x1 n1 x0 n0
#> 1 Study 1  0 76 12 42
#> 2 Study 2  0 13  2  2
#> 3 Study 3  0 17  2  3
#> 4 Study 4  5 47  7 36
#> 5 Study 5  0  4  3 33
#> 6 Study 6  0 50  0 76
#> 7 Study 7  0 41  0 37


# Transforming outcomes
dat2 <- data.frame(
  yk = c(dat$x1, dat$x0),
  zk = rep(1:0, each=7),
  nk = c(dat$n1, dat$n0),
  Study = c(dat$Study, dat$Study)
) %>%
  mutate(
    yk = yk / nk,
    vk = 1/(nk*yk) + 1/(nk-nk*yk)
  )
dat2
#>            yk zk nk   Study        vk
#> 1  0.00000000  1 76 Study 1       Inf
#> 2  0.00000000  1 13 Study 2       Inf
#> 3  0.00000000  1 17 Study 3       Inf
#> 4  0.10638298  1 47 Study 4 0.2238095
#> 5  0.00000000  1  4 Study 5       Inf
#> 6  0.00000000  1 50 Study 6       Inf
#> 7  0.00000000  1 41 Study 7       Inf
#> 8  0.28571429  0 42 Study 1 0.1166667
#> 9  1.00000000  0  2 Study 2       Inf
#> 10 0.66666667  0  3 Study 3 1.5000000
#> 11 0.19444444  0 36 Study 4 0.1773399
#> 12 0.09090909  0 33 Study 5 0.3666667
#> 13 0.00000000  0 76 Study 6       Inf
#> 14 0.00000000  0 37 Study 7       Inf
```


### Apply meta-analysis for GLMM

```{r example1-2, eval=FALSE}
formula <- yk ~ 1 + zk
dat3 <- dat2 %>% model.frame(formula=formula)

ma.grma <- metaGLMM(formula=formula, data=dat3, vi=dat2$vk, ni=dat2$nk, tau2=NA,
                family=binomial(link="logit"), tau2_var=TRUE)
summary(ma.grma)
#> Maximum likelihood estimation
#> 
#> Call:
#> bbmle::mle2(minuslogl = ll, start = c(beta_init, tau2 = 1), method = "L-BFGS-B", 
#>     skip.hessian = FALSE, lower = lower, upper = upper, control = list(maxit = 20000))
#> 
#> Coefficients:
#>             Estimate Std. Error z value   Pr(z)  
#> (Intercept)  -2.0223     1.2578 -1.6078 0.10788  
#> zk           -5.0609     2.6330 -1.9221 0.05459 .
#> tau2          9.2270     7.7225  1.1948 0.23216  
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> -2 log L: 173.5501



### confidence interval by profile likelihood
ci.pl <- confint_PL(ma.grma, parm="zk")
ci.pl
#>        lower     upper
#> zk -14.29106 -0.879445


### confidence interval by profile likelihood with simple Bartlett correction
ci.plsbc <- confint_SBC(ma.grma, parm="zk")
ci.plsbc
#>       lower        upper
#> zk -17.7262 -0.004056635


```



## Example 2: Poisson distributed outcome

### Data Preparation

```{r example2, eval=FALSE}
data(rutter2021, package="metaGLMM")
dat <- rutter2021
dat
#>                      Study Year    Sex Events   Time
#> 1         Watts et al 2009 2009 Female     13   9.04
#> 2        Dreyer et al 2011 2011 Female     16  23.75
#> 3        Birlik et al 2015 2015 Female     38  18.47
#> 4  Mohammed and Mandl 2015 2015 Female     13   9.38
#> 5  Romero-Gomez et al 2015 2015 Female      5   2.27
#> 6        Nesher et al 2016 2016 Female     11   2.72
#> 7       Saritas et al 2016 2016 Female     19   3.43
#> 8  Gudbrandsson et al 2017 2017 Female     68  26.00
#> 9         Makin et al 2017 2017 Female     13  15.48
#> 10         Park et al 2017 2017 Female    497 127.50
#> 11      Kanecki et al 2018 2018 Female    154  96.20
#> 12        Watts et al 2009 2009   Male      1   8.46
#> 13       Dreyer et al 2011 2011   Male      3  23.75
#> 14       Birlik et al 2015 2015   Male      3  18.47
#> 15 Mohammed and Mandl 2015 2015   Male      0   9.19
#> 16 Romero-Gomez et al 2015 2015   Male      0   2.27
#> 17       Nesher et al 2016 2016   Male      0   2.51
#> 18      Saritas et al 2016 2016   Male      4   3.34
#> 19 Gudbrandsson et al 2017 2017   Male     10  26.00
#> 20        Makin et al 2017 2017   Male      0  15.48
#> 21         Park et al 2017 2017   Male    115 127.50
#> 22      Kanecki et al 2018 2018   Male     23  96.20



# Standardize time variable
dat$Year <- dat$Year - min(dat$Year)


# Transforming outcomes
dat2 <- data.frame(
  yk = dat$Events,
  tk = dat$Time,
  xk = dat$Sex,
  year = dat$Year,
  Study = dat$Study
) %>%
  mutate(
    thetak = log((pmax(0.5, yk))/tk),
    yk = yk / tk,
    vk = 1 / pmax(0.5,dat$Events)
  )
dat2
#>           yk     tk     xk year                   Study     thetak          vk
#> 1  1.4380531   9.04 Female    0        Watts et al 2009  0.3632902 0.076923077
#> 2  0.6736842  23.75 Female    2       Dreyer et al 2011 -0.3949938 0.062500000
#> 3  2.0573904  18.47 Female    6       Birlik et al 2015  0.7214384 0.026315789
#> 4  1.3859275   9.38 Female    6 Mohammed and Mandl 2015  0.3263696 0.076923077
#> 5  2.2026432   2.27 Female    6 Romero-Gomez et al 2015  0.7896581 0.200000000
#> 6  4.0441176   2.72 Female    7       Nesher et al 2016  1.3972634 0.090909091
#> 7  5.5393586   3.43 Female    7      Saritas et al 2016  1.7118787 0.052631579
#> 8  2.6153846  26.00 Female    8 Gudbrandsson et al 2017  0.9614112 0.014705882
#> 9  0.8397933  15.48 Female    8        Makin et al 2017 -0.1745995 0.076923077
#> 10 3.8980392 127.50 Female    8         Park et al 2017  1.3604737 0.002012072
#> 11 1.6008316  96.20 Female    9      Kanecki et al 2018  0.4705232 0.006493506
#> 12 0.1182033   8.46   Male    0        Watts et al 2009 -2.1353492 1.000000000
#> 13 0.1263158  23.75   Male    2       Dreyer et al 2011 -2.0689702 0.333333333
#> 14 0.1624256  18.47   Male    6       Birlik et al 2015 -1.8175355 0.333333333
#> 15 0.0000000   9.19   Male    6 Mohammed and Mandl 2015 -2.9112631 2.000000000
#> 16 0.0000000   2.27   Male    6 Romero-Gomez et al 2015 -1.5129270 2.000000000
#> 17 0.0000000   2.51   Male    7       Nesher et al 2016 -1.6134299 2.000000000
#> 18 1.1976048   3.34   Male    7      Saritas et al 2016  0.1803236 0.250000000
#> 19 0.3846154  26.00   Male    8 Gudbrandsson et al 2017 -0.9555114 0.100000000
#> 20 0.0000000  15.48   Male    8        Makin et al 2017 -3.4326960 2.000000000
#> 21 0.9019608 127.50   Male    8         Park et al 2017 -0.1031842 0.008695652
#> 22 0.2390852  96.20   Male    9      Kanecki et al 2018 -1.4309351 0.043478261

```


### Apply meta-analysis for GLMM

```{r example2-2, eval=FALSE}

formula <- yk ~ 1 + xk + year + xk:year
dat3 <- dat2 %>% model.frame(formula=formula)


ma.grma <- metaGLMM(formula, data=dat3, vi=dat2$vk, ni=dat2$tk, tau2=NA, 
                family=poisson(link="log"), tau2_var=TRUE)
summary(ma.grma)
#> Maximum likelihood estimation
#> 
#> Call:
#> bbmle::mle2(minuslogl = ll, start = c(beta_init, tau2 = 1), method = "L-BFGS-B", 
#>     skip.hessian = FALSE, lower = lower, upper = upper, control = list(maxit = 20000))
#> 
#> Coefficients:
#>              Estimate Std. Error z value    Pr(z)   
#> (Intercept)  0.026855   0.516005  0.0520 0.958493   
#> xk          -2.587412   0.953510 -2.7136 0.006656 **
#> year         0.104032   0.077432  1.3435 0.179101   
#> xk:year      0.059344   0.135814  0.4369 0.662149   
#> tau2         0.381172   0.171376  2.2242 0.026136 * 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> -2 log L: 424.5847
```


### Bubble plot of meta-analysis for GLMM

```{r example2-3, eval=FALSE, fig.width=8, fig.height=6, fig.retina=2, out.width="80%", out.height="400px"}
g1 <- ggplot(dat2, aes(x=year, y=thetak)) +
  geom_point(aes(size=1/vk, fill=xk, shape=xk), color="black", alpha=0.7) +
  scale_size_continuous(range = c(2,10)) +
  scale_shape_manual(
    name   = "Sex",
    values = c("Female" = 21, "Male" = 22)
  ) +
  scale_fill_manual(
    name = "Sex",
    values = c("Female" = "pink", "Male" = "lightblue")
  ) +
  geom_abline(
    slope=coef(ma.grma)[3], 
    intercept=coef(ma.grma)[1], 
    color="red", 
    linewidth=1, linetype=1
  ) +
  geom_abline(
    slope=coef(ma.grma)[3]+coef(ma.grma)[4], 
    intercept=coef(ma.grma)[1]+coef(ma.grma)[2], 
    color="blue", 
    linewidth=1, linetype=1
  ) +
  guides(
    size = guide_legend(
      override.aes = list(
        fill = "white",
        color = "black",
        shape = 21
      )
    )
  ) +
  coord_cartesian(xlim = c(0,9)) +
  ylim(c(-4,3)) +
  labs(x = "Year", y = "log IR", size = "1/Variance") +
  theme(legend.position = "right")

g1
```


```{r example2-3-2, echo=FALSE, fig.width=8, fig.height=6, fig.retina=2, out.width="80%", out.height="400px"}
file <- if (knitr::is_latex_output()) {
  "figures/rda-rutter2021.pdf"
} else {
  "figures/rda-rutter2021.png"
}
knitr::include_graphics(file)
```



```{r example2-4, eval=FALSE}
### confidence interval by profile likelihood
ci.pl <- confint_PL(ma.grma)
ci.pl
#>                  lower      upper
#> (Intercept) -1.0570513  1.1062944
#> xk          -4.6276377 -0.7489715
#> year        -0.0576682  0.2655337
#> xk:year     -0.2119976  0.3433750



### confidence interval by profile likelihood with simple Bartlett correction
ci.plsbc <- confint_SBC(ma.grma)
ci.plsbc
#>                   lower      upper
#> (Intercept) -1.12468415  1.1735710
#> xk          -4.75295689 -0.6445385
#> year        -0.06780246  0.2756033
#> xk:year     -0.22810411  0.3605470
```



## Example 3: Gamma distributed outcome

### Data Preparation

```{r example3, eval=FALSE}
data(long2020, package="metaGLMM")

# Transforming outcomes
dat <- long2020 %>% mutate(
  yi = log(mi),
  vi = sdi^2 / mi^2
)

head(dat)
#>     Study   mi sdi ni trt       yi         vi
#> 1 Study 1  9.6 0.7 20   1 2.261763 0.00531684
#> 2 Study 2  9.9 8.3 25   1 2.292535 0.70288746
#> 3 Study 3 13.8 4.2 23   1 2.624669 0.09262760
#> 4 Study 4 16.5 7.4 18   1 2.803360 0.20113866
#> 5 Study 5  8.2 4.3 75   1 2.104134 0.27498513
#> 6 Study 1 14.6 2.2 20   0 2.681022 0.02270595
```



### Apply meta-analysis for GLMM

```{r example3-2, eval=FALSE}

formula <- mi ~ 1 + trt
dat2 <- dat %>% model.frame(formula=formula)


ma.grma <- metaGLMM(formula, data=dat2, vi=dat$vi, ni=dat$ni, tau2=NA, 
                family=Gamma(link = "log"), tau2_var=TRUE)
summary(ma.grma)
#> Maximum likelihood estimation
#> 
#> Call:
#> bbmle::mle2(minuslogl = ll, start = c(beta_init, tau2 = 1), method = "L-BFGS-B", 
#>     skip.hessian = FALSE, lower = lower, upper = upper, control = list(maxit = 20000))
#> 
#> Coefficients:
#>             Estimate Std. Error z value   Pr(z)    
#> (Intercept)  2.85140    0.11597 24.5874 < 2e-16 ***
#> trt         -0.43137    0.15921 -2.7094 0.00674 ** 
#> tau2         0.06796    0.03587  1.8946 0.05814 .  
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> -2 log L: 50226.91



### confidence interval by profile likelihood
ci.pl <- confint_PL(ma.grma, parm="trt")
ci.pl
#>         lower       upper
#> trt -0.681928 -0.04249324



### confidence interval by profile likelihood with simple Bartlett correction
ci.plsbc <- confint_SBC(ma.grma, parm="trt")
ci.plsbc
#>          lower      upper
#> trt -0.8958982 0.03439087
```




## Citetion







